Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-20 RevisedSubmission/hines_hetland_2017_gji.tex
add shit

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Moritz1978}. The latter has been used by \citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain rates from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing discrete GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a continuous stochastic process whose value at any collection of points can be described as a Gaussian random vector. Just as a Gaussian random vector is fully determined by a mean vector and a covariance matrix, a Gaussian process is fully determined by a mean function and a covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has a mean function $\mathrm{E}[B(t)] = 0$ and a covariance function $\mathrm{Cov}\left[B(t),B(t')\right] = \min(t,t')$, where $t,t' \ge 0$. If there is no information available to help choose an appropriate prior Gaussian process, then maximum likelihood methods can be used to objectively choose one that is most consistent with the data. Once a prior is selected, we incorporate GNSS observations with the prior to form a posterior estimate of transient strain. The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. We use GPR to infer strain resulting from SSEs in the Pacific Northwest, demonstrating that GPR is an effective tool for detecting transient geophysical processes.

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p) = (u_\mathrm{e},u_\mathrm{n})$.

We make a prior assumption that each component of $\vec{u}$ is a Gaussian process with zero mean and a generic covariance function $\mathrm{Cov}\left[u_i(p),u_i(p')\right] = C_{u_i}(p,p')$. Using a more concise notation, we can equivalently state our prior on each component of $\vec{u}$ as $u_i \sim \mathcal{GP}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}(p,p') = C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = \phi^2 X_i(\vec{x},\vec{x}')T_i(t,t'),
\end{equation}   
where the parameter $\phi$ controls the amplitude of realizations of $u_i$. As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. We also require that the derivatives $\partial^2 X_i(\vec{x},\vec{x}')/ \partial x_j \partial x_j'$ and $\partial^2 T_i(t,t') / \partial t \partial t'$ exist. This ensures that $u_i$ is spatially and temporally differentiable, allowing us to compute $\dot\varepsilon$ (See \citet[sec. 2.2]{Adler1981} or \citet[sec. 10A]{Papoulis1991} for a definition of stochastic differentiation and the conditions for differentiability).

We now provide an example to give our prior on $\vec{u}$ a more tangible meaning. We can satisfy our requirements for positive definiteness and differentiability by using a squared exponential function for $X_i$ and $T_i$,
\begin{equation}\label{eq:ExampleCov}
X_i(\vec{x},\vec{x}') = \exp\left(\frac{-||\vec{x} - \vec{x}'||_2^2}{2\ell^2}\right), \ \ \ \ T_i(t,t') = \exp\left(\frac{-|t - t'|^2}{2\tau^2}\right).
\end{equation}
The parameters $\ell$ and $\tau$ control the length-scale and time-scale, respectively, of realizations of $u_i$. An example realization is shown in Figure ...

While the squared exponential function is a commonly used covariance function for GPR, it is not appropriate for every application. The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe. To keep this section sufficiently general, we hold off on specifying $X_i$ and $T_i$ until Section \ref{sec:SignalModel}, where we estimate transient strain from slow slip events in the Pacific Northwest.

@@

                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).

@@

Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian processes $\hat{u}_i$ and $\hat{u}_j$, we know that $\dot\varepsilon_{ij}$ is also a Gaussian process. The mean and covariance functions for $\dot{\varepsilon}_{ij}$ are given by
\begin{equation}\label{eq:StrainMean}
\mu_{\dot\varepsilon_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)

@@

\begin{equation}\label{eq:StrainCov}
C_{\dot\varepsilon_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right),
respectively \citep[sec. 10]{Papoulis1991}.

>>>>

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Moritz1978}. The latter has been used by \citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain rates from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing discrete GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a continuous stochastic process whose value at any collection of points can be described as a Gaussian random vector. Just as a Gaussian random vector is fully determined by a mean vector and a covariance matrix, a Gaussian process is fully determined by a mean function and a covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has the mean function $\mathrm{E}[B(t)] = 0$ and the covariance function $\mathrm{Cov}\left[B(t),B(t')\right] = \min(t,t')$, where $t,t' \ge 0$. If there is no information available to help choose an appropriate prior Gaussian process, then maximum likelihood methods can be used to objectively choose one that is most consistent with the data. Once a prior is selected, we incorporate GNSS observations with the prior to form a posterior estimate of transient strain. The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. We use GPR to infer strain resulting from SSEs in the Pacific Northwest, demonstrating that GPR is an effective tool for detecting transient geophysical processes.

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, which are constrained by GNSS data.

We let $\vec{u}(p) = (u_\mathrm{e}(p),u_\mathrm{n}(p))$ be our prior understanding of transient displacements. Since transient displacements are not precisely known, we do not consider $\vec{u}$ to be a deterministic function. Instead, $\vec{u}$ is a stochastic process containing a distribution of functions that could potentially describe transient displacements. We let each component of $\vec{u}$ be a Gaussian process with zero mean and a generic covariance function $\mathrm{Cov}\left[u_i(p),u_i(p')\right] = C_{u_i}(p,p')$. Using a more concise notation, we write our prior on each component of $\vec{u}$ as $u_i \sim \mathcal{GP}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite in order to be a valid covariance function. By definition, $C_{u_i}$ is a positive definite function if the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}(p,p') = C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}   
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite \citep[sec. 4.2.4]{Rasmussen2006}. We also require that the derivatives $\partial^2 X_i(\vec{x},\vec{x}')/ \partial x_j \partial x_j'$ and $\partial^2 T_i(t,t') / \partial t \partial t'$ exist. This ensures that $\vec{u}$ is spatially and temporally differentiable, allowing us to compute $\dot\varepsilon$ (See \citet[sec. 2.2]{Adler1981} or \citet[sec. 10A]{Papoulis1991} for a definition of stochastic differentiation and the conditions for differentiability).

To give an example of $u_i$, we can use a squared exponential function for $X_i$ and $T_i$,
\begin{equation}\label{eq:ExampleCov}
X_i(\vec{x},\vec{x}') = \exp\left(\frac{-||\vec{x} - \vec{x}'||_2^2}{2\ell^2}\right), \ \ \ \ T_i(t,t') = \phi^2 \exp\left(\frac{-|t - t'|^2}{2\tau^2}\right),
\end{equation}
which satisfies our requirements for positive definiteness and differentiability. The parameters $\ell$ and $\tau$ control the length-scale and time-scale, respectively, of realizations (i.e., sample functions) of $u_i$. The parameter $\phi$, which we have arbitrarily chosen to incorporate into $T_i$ rather than $X_i$, controls the amplitude of realizations of $u_i$. Ideally, we want realizations of $u_i$ to have a length-scale, time-scale, and amplitude that resembles what we expect from true signal. A few realizations for $u_i$ are shown in Figure XXX. While the squared exponential function is a commonly used covariance function for GPR, it is not appropriate for every application. The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe. To keep this section sufficiently general, we hold off on specifying $X_i$ and $T_i$ until Section \ref{sec:SignalModel}, where we estimate transient strain from slow slip events in the Pacific Northwest.

@@

                                        \frac{\partial \hat{u}_j(p)}{\partial x_i} \right).

@@

Since eq. (\ref{eq:StrainRate}) is a linear operation on the Gaussian processes $\hat{u}_i$ and $\hat{u}_j$, we know that $\dot\varepsilon_{ij}$ is also a Gaussian process. From \citet[sec. 10]{Papoulis1991}, we find that the mean and covariance functions for $\dot{\varepsilon}_{ij}$ are 
\begin{align}\label{eq:StrainMean}
\mu_{\dot\varepsilon_\mathrm{ee}}(p) &= \frac{\partial^2 \mu_{\hat{u}_\mathrm{e}}(p)}{\partial t \, \partial x_\mathrm{e}} \\
\mu_{\dot\varepsilon_\mathrm{nn}}(p) &= \frac{\partial^2 \mu_{\hat{u}_\mathrm{n}}(p)}{\partial t \, \partial x_\mathrm{n}} \\
\mu_{\dot\varepsilon_\mathrm{en}}(p) &= \frac{1}{2}\frac{\partial}{\partial t}\left(
                                        \frac{\partial \mu_{\hat{u}_\mathrm{e}}(p)}{\partial x_\mathrm{n}} + 
                                        \frac{\partial \mu_{\hat{u}_\mathrm{n}}(p)}{\partial x_\mathrm{e}} \right)                                  

@@

\begin{align}\label{eq:StrainCov}
C_{\dot\varepsilon_\mathrm{ee}}(p,p') &= \frac{\partial^4 C_{\hat{u}_\mathrm{e}}(p,p')}{\partial t \, \partial t' \, \partial x_\mathrm{e} \, \partial x'_\mathrm{e}} \\
C_{\dot\varepsilon_\mathrm{nn}}(p,p') &= \frac{\partial^4 C_{\hat{u}_\mathrm{n}}(p,p')}{\partial t \, \partial t' \, \partial x_\mathrm{n} \, \partial x'_\mathrm{n}} \\
C_{\dot\varepsilon_\mathrm{en}}(p,p') &= \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                         \frac{\partial^2 C_{\hat{u}_\mathrm{e}}(p,p')}{\partial x_\mathrm{n} \, \partial x'_\mathrm{n}} + 
                                         \frac{\partial^2 C_{\hat{u}_\mathrm{n}}(p,p')}{\partial x_\mathrm{e} \, \partial x'_\mathrm{e}} \right),
respectively.
